---
title: "ANOVA"
format: html
editor: visual
---

##                                     One-Way ANOVA

ANOVA (Analysis of Variance) is mainly used when your independent variable(s) are categorical and your dependent variable is continuous (numerical).

So, in short:

Independent variable(s): Categorical (e.g., gender, treatment type, region, etc.)

Dependent variable: Continuous (e.g., income, test scores, height, etc.)

### What does ANOVA test?

ANOVA tests whether the means of the dependent variable differ significantly across the levels (groups) of the categorical variable.

Example:

Suppose you want to know if average exam scores differ across 3 teaching methods (Method A, Method B, Method C).

**Teaching method** → categorical (independent variable)

**Exam score** → continuous (dependent variable)

✔️ ANOVA is suitable here.

### ASSUMPTIONS

**Normality** - each sample is taken from a normally distributed population. **Independance** - each sample has been drawn independently of the other samples. **Equal variance** - variance of data in the different groups should be the same.

```{r}
 library(dplyr)
library(tidyverse)

 data(iris)
 dat <- iris %>%
 select(Species, Sepal.Length)
 summary(dat)
```

### 

```{r}
iris %>% 
  group_by(Species) %>% 
  summarize (count_species =n(),
             mean_speal_length =mean(Sepal.Length,na.rm=TRUE),
             sd_sepal_length =sd(Sepal.Length,na.rm = TRUE)
             )
```

```{r}
ggplot(iris, aes(x = Species, y = Sepal.Length)) + 
geom_boxplot()
```

```{r}
one_way_anova <- aov(Sepal.Length ~ Species , data = iris)
 summary(one_way_anova)
```

P −Value \< 0.05. So, we reject H0. We can conclude that, there is
a statistical difference between the mean of the sepal length
according to the type of the iris plant at 5% level of significance.

#### Multiple Comparisons (POST-HOC TESTS)

If the test rejects H0 , the one-way ANOVA test does not inform
which group has a different mean. Then run a pairwise t-test for
multiple comparison and identify the different groups.

```{r}
pairwise.t.test(iris$Sepal.Length,iris$Species,p.adj="bonferroni")
```

-   Based on P-values \< 0.05, all the three species combinations are
    differ according to the Sepal Length.

-    There are several possible adjustment methods: “**bonferroni**”,
    “**holm**”, “**hochberg**”, “**hommel**”, “**BH**”

-   Tukey’s Honest Significant Differences (Tukey’s HSD) is also a
    very common test that we can use for multiple comparisons. It
    requires an anova object.

    ```{r}
    pairwise.t.test(iris$Sepal.Length,iris$Species,p.adj="holm")
    ```

    ```{r}
    TukeyHSD(one_way_anova)
    ```

### Check Normality Assumptions

### **Normality**

1.  **Analyzing the ANOVA model residuals** to check the normality
    for all groups together. This approach is easier and it’s very handy
    when you have many groups or if there are few data points per
    group.

    Normality of residuals can be tested visually via a QQ-plot, and/or
    formally via a normality test (Shapiro-Wilk test).

    ```{r}
    library(car)
    qqPlot(one_way_anova$residuals,id=FALSE)
    ```

H0 : data come from a normal distribution

H1 : data do not come from a normal distribution.

```{r}
shapiro.test(one_way_anova$residuals)

# p value > 0.05 and we do not reject H0 and conclude that the residulas follow approximately a normal distibution at 5% level of sgnificance.

```

1.  **Check normality for each group separately**. This approach might
    be used when you have only a few groups and many data points
    per group.

    ```{r}
    iris %>% 
      group_by(Species) %>%
      summarize(statistic =shapiro.test(Sepal.Length)$statistic,
                p.value =shapiro.test(Sepal.Length)$p.value
                )
    ```

### HOMOGENEITY OF VARIANCES

*The result will have an impact on whether we use the ANOVA
or the Welch ANOVA.*

We can formally test for equality of the variances with a
**Levene’s** or **Bartle ’s** test. Notice that the **Levene’s test** is less
sensitive to departures from normal distribution than the
**Bartle ’s test**.

H0 : Variance are equal

H1 : at least one Variance is different

```{r}
bartlett.test(Sepal.Length~Species,data = iris)
```

```{r}
leveneTest(Sepal.Length~Species,data = iris)
```

**Welch One-Way ANOVA**

The Welch One ANOVA test is an alternative to the standard one way ANOVA in the situation where the homogeneity of variance cant be assumed.

```{r}
library(rstatix)

waov <- iris %>%
  welch_anova_test(Sepal.Length ~ Species)
waov
  
```

In this case ,the **Games -Howell post hoc test** or **Paiwise t-test** can be used to compare all possible combinations of group differences.

```{r}
pwc2 <- iris %>% games_howell_test(Sepal.Length ~ Species)
pwc2
```

```{r}
pwc3 <- iris %>%
  pairwise_t_test(
    Sepal.Length ~Species ,pool.sd = FALSE,
    p.adjust.method = "bonferroni"
  )
pwc3
```

##                                    Two-Way ANOVA

Two-Way ANOVA is an extension of One-Way ANOVA, used to examine the effect of two independent categorical variables on one continuous dependent variable — and also their interaction effect.

When to use Two-Way ANOVA:

-   You have two categorical independent variables (factors)

-   You have one continuous dependent variable.

-   You want to test:

1.  Main effect of Factor A

2.  Main effect of Factor B

3.  Interaction effect (A × B)
